Functional Programming in Computer Science

### 2015-12-18 loren/SULI/finalReport.tex
final drafts

Here, the expression \texttt{$\lambda$x ->} signifies to substitute \texttt{3} in for \texttt{x}. Although this substitution seemed easy, the general process is much more difficult in pure lambda calculus where predefined constants and functions are not allowed. In pure lambda calculus, lambda expressions are evaluated through a process called reduction. This is primarily through substituting a free variable for an expression in a lambda expression.  In general, an expression \texttt{$(\lambda$ x -> E) E1} is reduced by substituting \texttt{E1} in for all variables \texttt{x} in \texttt{E}.

@@

If in fact a free variable becomes bound as a result of the substitution, the reduction involved \emph{variable capture}. Variable capture is one of the main difficulties of reducing lambda expressions. The state of variable capture is shown below with the generic addition function.

@@

Alpha reduction allows one to change the bound variables of a lambda expression barring variable capture. Note that conversion does not change the lambda expression; both the original and new lambda expressions are equivalent. An example alpha conversion from the previous example is shown below.

@@

Due to its frequent appearance, \texttt{($\lambda$v -> E) E1} is called a $\beta$-redex or ``reduction expression". Beta reduction is the most involved rule of evaluation in lambda calculus.

@@

We want a methodical approach to choose which beta redexes to reduce first to get to normal form. First, not every lambda expression can be reduced to a normal form due to the classic example illustrated below.

@@

For any lambda expression of the form \emph{\texttt{E = (($\lambda$x -> B) A)}}, we say that beta redex \emph{\texttt{E}} is outside any beta redex that occurs in \emph{\texttt{B}} or \emph{\texttt{A}} and that these are inside \emph{\texttt{E}}. A $\beta$-redex \emph{\texttt{E}} is outside any $\beta$-redex that occurs in \emph{\texttt{B}} or \emph{\texttt{A}} and that these are inside \emph{\texttt{E}}. A $\beta$-redex in a lambda expression is outermost if there is no $\beta$-redex outside of it, and it is innermost if there is no $\beta$-redex inside of it.
\end{definition}

@@

The example  below distinguish the differences between outermost and innermost reductions.

\begin{center}
\texttt{($\lambda$x -> ($\lambda$y -> (add x (($\lambda$x -> (sub x 3)) y)))) 5 6}\parskip 0pt
\end{center}

\begin{tabular}{l | l}
\multicolumn{1}{c}{ \underline{Normal Order Reduction}} & \multicolumn{1}{c}{ \underline{Applcative Order Reduction}}\\
\texttt{= ($\lambda$y -> (add 5 (($\lambda$x -> (sub x 3)) y))) 6}&\texttt{= ($\lambda$x -> ($\lambda$y -> (add x ((sub y 3))))) 5 6}\\
\texttt{= (add 5 (($\lambda$x -> (sub x 3)) 6))}&\texttt{= ($\lambda$y -> (add 5 ((sub y 3)))) 6}\\
\texttt{= (add 5 ((sub 6 3)))}&\texttt{= (add 5 ((sub 6 3)))}\\
\texttt{= (add 5 3)}&\texttt{= (add 5 3)}\\
\texttt{= 8}&\texttt{= 8}\\
\end{tabular}

@@

Church's Thesis states that lambda calculus is Turing complete. Motivated by this fact, we demonstrate how to define nonnegative integers, basic mathematical functions, logical operators, list constructors, and how to simulate recursion.

@@

Other arithmetic operations such as subtract, multiply, exponentiate. The next set of lambda expressions is the set of arithmetic operators. The boolean values true and false are the foundations of the logical operators and many other functions in general. As in the case of defining the nonnegative integers, the stated definitions below of true and false are not the only way to define true, but they may be the most efficient. The operators and, or, and not are also defined below, and they are built from the definitions of true and false.

@@

Conditional statements are left to the appendix. It is possible to contruct lists of arbitrary sizes by first constructing a \texttt{pair} function and then continually nesting \texttt{pair} functions in the second slot to the desired length (see appendix). Finally, the most efficient way to simulate recursion is through the fixed point combinator \texttt{Y}.

@@

As functions do not have names in pure lambda calculus, it is impossible for a function to achieve recursion through calling itself. Luckily, the fixed point combinator acts in the following way:

@@

True to its name, \texttt{f} is a fixed point of \texttt{Y}. The key to exploiting the property of the fixed-point combinator is to clevely set up a conditional statement within the function. The pseudocode of the \texttt{factorial} function is displayed below with a small evaluation in impure lambda calculus.

\small
\hspace{2cm}\texttt{fac = ($\lambda$m -> ($\lambda$n -> (if n = 0 then 1 else n $\times$ (m (n - 1)))))}

\hspace{2cm}\texttt{Y fac 3}

\hspace{2.5cm}\texttt{= fac (Y fac) 3}

\hspace{2.5cm}\texttt{= ($\lambda$m -> ($\lambda$n -> (if n = 0 then 1 else n $\times$ (m (n - 1))))) (Y fac) 3}

\hspace{2.5cm}\texttt{= (if 3 = 0 then 1 else 3 $\times$ ((Y fac) (3 - 1)))}

\hspace{2.5cm}\texttt{= 3 $\times$ ((Y fac) 2)}

\hspace{2.5cm}\texttt{= 3 $\times$ (fac (Y fac) 2)}

\hspace{2.5cm}\texttt{= 3 $\times$ (($\lambda$m -> ($\lambda$n -> (if n = 0 then 1 else n $\times$ (m (n - 1))))) (Y fac) 2)}

\hspace{2.5cm}\texttt{= 3 $\times$ ((if 2 = 0 then 1 else 2 $\times$ ((Y fac) (2 - 1))))}

\hspace{2.5cm}\texttt{= 3 $\times$ (2 $\times$ ((Y fac) 1))}

\hspace{2.5cm}\texttt{= 3 $\times$ (2 $\times$ (fac (Y fac) 1))}

\hspace{2.5cm}\texttt{= 3 $\times$ (2 $\times$ (($\lambda$m -> ($\lambda$n -> (if n = 0 then 1 else n $\times$ (m (n - 1))))) (Y fac) 1))}

\hspace{2.5cm}\texttt{= 3 $\times$ (2 $\times$ (if 1 = 0 then 1 else 1 $\times$ ((Y fac) (1 - 1))))}

\hspace{2.5cm}\texttt{= 3 $\times$ (2 $\times$ (1 $\times$ ((Y fac) 0))}

\hspace{2.5cm}\texttt{= 3 $\times$ (2 $\times$ (1 $\times$ (fac (Y fac) 0))}

\hspace{2.5cm}\texttt{= 3 $\times$ (2 $\times$ (1 $\times$ (($\lambda$m -> ($\lambda$n -> (if n = 0 then 1 else n $\times$ (m (n - 1)))))(Y fac) 0))}

\hspace{2.5cm}\texttt{= 3 $\times$ (2 $\times$ (1 $\times$ (if 0 = 0 then 1 else 0 $\times$ ((Y fac) (0 - 1)))))}

\hspace{2.5cm}\texttt{= 3 $\times$ (2 $\times$ (1 $\times$ 1)) = 6}
\normalsize

@@

My ultimate task was to create a lambda parser and evaluator in Haskell. To
do this, I first read """""""""""' and completed all exercises.  The
content of the material was similar to the content presented here in the
previous sections. Many exercises were detailed and required $\beta$
reduction to be completed by hand. Following, I created a lambda parser and
evaluator using both normal order and applicative order evaluation. In
order to complete the variable substitution with new variables, it was
necessary to provide a fresh variable supply. The first implementation
involved creating an infinite list of variables, exploiting Haskell's lazy
evaluation. This list could be split in half when a substitution needed to
be made in a combined expression (case d) in definition 4. The easiest way
to split the list in half was to create two new lists by appending every
other element to the end of each list. Since only finitely many fresh
variables were needed, starting with an infinite list of fresh variables
was sufficient. The second implementation was to thread the variables
through the program. After a expression was reduced, the fresh variable
supply was passed to inner expressions to complete any reductions. Once all
inner expressions were reduced, the remaining fresh variable supply was
passed to another expression until the entire expression was in normal
form. This method is displayed in appendix .... After creating the parser
and evaluator, I constructed the nonnegative integers, arithmetic
expressions, logical operators, and list operations. The final exercise was
to code enough functions to parse and evaluate the factorial function in
lambda calculus, as shown above.

>>>>

Here, the expression \texttt{$\lambda$x ->} signifies to substitute \texttt{3} in for \texttt{x}. Although this substitution seemed easy, the general process is much more difficult in pure lambda calculus where predefined constants and functions are not allowed. In pure lambda calculus, lambda expressions are evaluated through a process called reduction. In general, an expression \texttt{$(\lambda$x -> E) E1} is reduced by substituting \texttt{E1} in for all free variables \texttt{x} in \texttt{E}.

@@

If in fact a free variable becomes bound as a result of the substitution, the reduction involved variable capture; variable capture is one of the main difficulties of reducing lambda expressions. The state of variable capture is shown below with the generic addition function.

@@

Alpha reduction allows one to change the bound variables of a lambda expression barring variable capture. Note that conversion does not change the lambda expression: both the original and new lambda expressions are equivalent. An example alpha conversion from the previous example is shown below.

@@

Due to its frequent appearance, \texttt{($\lambda$v -> E) E1} is called a $\beta$-redex or ``reduction expression." Beta reduction is the most involved rule of evaluation in lambda calculus.

@@

We want a methodical approach to choose which $\beta$-redexes to reduce first to get to normal form. First, not every lambda expression can be reduced to a normal form due to the classic example illustrated below.

@@

For any lambda expression of the form \emph{\texttt{E = (($\lambda$x -> B) A)}}, we say that $\beta$-redex \emph{\texttt{E}} is outside any $\beta$-redex that occurs in \emph{\texttt{B}} or \emph{\texttt{A}} and that these are inside \emph{\texttt{E}}. A $\beta$-redex \emph{\texttt{E}} is outside any $\beta$-redex that occurs in \emph{\texttt{B}} or \emph{\texttt{A}} and that these are inside \emph{\texttt{E}}. A $\beta$-redex in a lambda expression is outermost if there is no $\beta$-redex outside of it, and it is innermost if there is no $\beta$-redex inside of it.
\end{definition}

@@

The example  below distinguishes the differences between outermost and innermost reductions.

\begin{center}
\texttt{($\lambda$x -> ($\lambda$y -> (add x (($\lambda$x -> (sub x 3)) y)))) 5 6}
\end{center}

\begin{tabular}{l | l}
\multicolumn{1}{c}{ \underline{Normal Order Reduction}} & \multicolumn{1}{c}{ \underline{Applcative Order Reduction}}\\
\texttt{=($\lambda$y -> (add 5 (($\lambda$x ->(sub x 3))y))) 6}&\texttt{=($\lambda$x -> ($\lambda$y -> (add x((sub y 3))))) 5 6}\\
\texttt{=(add 5 (($\lambda$x -> (sub x 3)) 6))}&\texttt{=($\lambda$y -> (add 5 ((sub y 3)))) 6}\\
\texttt{=(add 5 ((sub 6 3)))}&\texttt{=(add 5 ((sub 6 3)))}\\
\texttt{=(add 5 3)}&\texttt{=(add 5 3)}\\
\texttt{=8}&\texttt{=8}\\
\end{tabular}

@@

Church's Thesis states that lambda calculus is Turing-complete. Motivated by this fact, we demonstrate how to define nonnegative integers, basic mathematical functions, logical operators, list constructors, and how to simulate recursion.

@@

Other arithmetic operations implemented include \texttt{subtract}, \texttt{multiply}, and \texttt{exponentiate}. The next set of lambda expressions is the set of logical operators. The boolean values \texttt{true} and \texttt{false} are the foundations of the logical operators and many other functions in general. As in the case of defining the nonnegative integers, the stated definitions below of \texttt{true} and \texttt{false} are not the only definitions, but they may be the most efficient. The operators \texttt{and}, \texttt{or}, and \texttt{not} are also defined below, and they are built from the definitions of \texttt{true} and \texttt{false}.

@@

Conditional statements are left to the appendix. It is possible to contruct lists of arbitrary sizes by first constructing a \texttt{pair} function and then continually nesting \texttt{pair} functions in the second slot to the desired length (see appendix 7.3). Finally, the most efficient way to simulate recursion is through the fixed point combinator \texttt{Y}.

@@

As functions do not have names in pure lambda calculus, it is impossible for a function to achieve recursion through calling itself. Luckily, the fixed point combinator acts as follows:

@@

True to its name, \texttt{f} is a fixed point of \texttt{Y}. The key to exploiting the property of the fixed-point combinator is to clevely set up a conditional statement within the input function. The pseudocode of the \texttt{factorial} function is displayed below with a small evaluation in impure lambda calculus.

\small
\hspace{0.5cm}\texttt{fac = ($\lambda$m -> ($\lambda$n -> (if n = 0 then 1 else n $\times$ (m (n - 1)))))}

\hspace{0.5cm}\texttt{Y fac 3}

\hspace{1cm}\texttt{= fac (Y fac) 3}

\hspace{1cm}\texttt{= ($\lambda$m -> ($\lambda$n -> (if n = 0 then 1 else n $\times$ (m (n - 1))))) (Y fac) 3}

\hspace{1cm}\texttt{= (if 3 = 0 then 1 else 3 $\times$ ((Y fac) (3 - 1)))}

\hspace{1cm}\texttt{= 3 $\times$ ((Y fac) 2)}

\hspace{1cm}\texttt{= 3 $\times$ (fac (Y fac) 2)}

\hspace{1cm}\texttt{= 3 $\times$ (($\lambda$m -> ($\lambda$n -> (if n = 0 then 1 else n $\times$ (m (n - 1))))) (Y fac) 2)}

\hspace{1cm}\texttt{= 3 $\times$ ((if 2 = 0 then 1 else 2 $\times$ ((Y fac) (2 - 1))))}

\hspace{1cm}\texttt{= 3 $\times$ (2 $\times$ ((Y fac) 1))}

\hspace{1cm}\texttt{= 3 $\times$ (2 $\times$ (fac (Y fac) 1))}

\hspace{1cm}\texttt{= 3 $\times$ (2 $\times$ (($\lambda$m -> ($\lambda$n -> (if n = 0 then 1 else n $\times$ (m (n - 1))))) (Y fac) 1))}

\hspace{1cm}\texttt{= 3 $\times$ (2 $\times$ (if 1 = 0 then 1 else 1 $\times$ ((Y fac) (1 - 1))))}

\hspace{1cm}\texttt{= 3 $\times$ (2 $\times$ (1 $\times$ ((Y fac) 0))}

\hspace{1cm}\texttt{= 3 $\times$ (2 $\times$ (1 $\times$ (fac (Y fac) 0))}

\hspace{1cm}\texttt{= 3 $\times$ (2 $\times$ (1 $\times$ (($\lambda$m -> ($\lambda$n -> (if n = 0 then 1 else n $\times$ (m (n - 1)))))(Y fac) 0))}

\hspace{1cm}\texttt{= 3 $\times$ (2 $\times$ (1 $\times$ (if 0 = 0 then 1 else 0 $\times$ ((Y fac) (0 - 1)))))}

\hspace{1cm}\texttt{= 3 $\times$ (2 $\times$ (1 $\times$ 1)) = 6}
\normalsize

@@

My overall task was to create a lambda parser and evaluator in Haskell. To
do this, I first read \textit{Formal Syntax and Semantics of Programming Languages} by Kurtz and Slonneger and completed all exercises.$^5$  The
content of the material was similar to the content presented here in the
previous sections. Many exercises were detailed and required beta
reduction to be completed by hand. Following, I created a lambda parser and
evaluator using both normal order and applicative order evaluation. In
order to complete the variable substitution with new variables, it was
necessary to provide a fresh variable supply. The first implementation
involved creating an infinite list of variables, exploiting Haskell's lazy
evaluation. This list could be split in half when a substitution needed to
be made in a combined expression (case d in Definition 4). The easiest way
to split the list in half was to create two new lists by appending every
other element to the end of each list. Since only finitely many fresh
variables were needed, starting with an infinite list of fresh variables
was sufficient. The second implementation was to thread the variables
through the expression. After a expression was reduced, the fresh variable
supply was passed to inner expressions to complete any reductions. Once all
inner expressions were reduced, the remaining fresh variable supply was
passed to another expression until the entire expression was in normal
form. This method is displayed in appendix 7.3. After creating the parser
and evaluator, I constructed the nonnegative integers, arithmetic
expressions, logical operators, and list operations. The final exercise was
to code enough functions to parse and evaluate the \texttt{factorial} function in pure
lambda calculus.
